Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-12 submission/manuscript.Rmd
edit intro and add females

A single genome holds the instructions to generate the myriad of cell types found within the adult organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. Many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Recent advancements in next generation sequencing unexpectedly revealed mutations in many chromatin regulators cause neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018]. Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. Very few studies have investigated this cellular identity crisis in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if ectopic gene expression contributes to neurodevelopmental impairments.

To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the nature of genes dysregulated and the molecular mechanisms governing their de-repression. In this study, we characterized the aberrant expression of tissue-enriched genes with loss of lysine demethylase 5C (KDM5C). KDM5C, , also known as SMCX or JARID1C, is a chromatin regulator that can repress gene expression through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], marks enriched at active gene promoters. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose predominant features include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017]. Unexpectedly, RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017]. It is currently unknown what types of testis genes are dysregulated, when in _Kdm5c_-KO development this begins, and if other tissue-specific genes are also aberrantly expressed with KDM5C loss.

The testis contains both germ cells (cells that pass on their genetic material, e.g. sperm) and somatic cells (cells that perform all other bodily functions). Distinguishing between the germline and the soma is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] and that typically occurs during early embryogenesis[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency. Initially, repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], are placed at germline gene promoters in embryonic stem cells and are then decorated with DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] in the post-implantation embryo. How KDM5C interacts with known germline gene silencers during this time point is currently unknown. Additionally, most studies have focused on select genes important for early germ cell development rather than germline genes as a whole given the lack of a curated list of germline-enriched genes. Thus, it is unclear if the mechanism of repression differs for certain classes of germline genes, e.g. meiotic genes versus spermatid differentiation genes. Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.

- Add females - KDM5C is a sexually dimorphic chromatin regulator, embryonic lethality makes it difficult to compare

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including testis, liver, muscle, and ovary-enriched genes. Both the _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes. We demonstrated testis-enriched genes genes are germline genes and not somatic testis genes by. We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis. Additionally, KDM5C was bound to only a subset of germline genes expressed in _Kdm5c_-KO EpiLCs, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms. Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA CpG methylation in EpiLCs. Thus, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

@@

Previous RNA sequencing (RNA-seq) in the adult hippocampus revealed ectopic expression of some testis genes within the _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017]. Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, if the testis is the only tissue type misexpressed. Therefore, we systematically assessed the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the adult amygdala and hippocampus in mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].

>>>>

A single genome holds the instructions to generate the myriad of cell types found within the adult organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. Many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Unexpectedly, next generation sequencing revealed mutations in many chromatin regulators directly cause or are associated with neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018]. Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. Very few studies have investigated this cellular identity crisis in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if ectopic gene expression contributes to neurodevelopmental impairments.

To elucidate the role of tissue identity in chromatin-linked neurodevelopmental disorders, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression. In this study, we characterized the aberrant expression of tissue-enriched genes with loss of lysine demethylase 5C (KDM5C). KDM5C, also known as SMCX or JARID1C, is a chromatin regulator that can repress gene expression through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], marks enriched at active gene promoters. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and often more severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017]. Unexpectedly, RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus revealed ectopic expression of testis genes within the brain[@scandagliaLossKdm5cCauses2017]. It is currently unknown what types of testis genes are dysregulated, when in _Kdm5c_-KO development this begins, if this phenomenon is sexually dimorphic, and if other tissue-specific genes are also aberrantly expressed with KDM5C loss.

The testis contains both germ cells (cells that pass on their genetic material, e.g. sperm) and somatic cells (cells that perform all other bodily functions). Distinguishing between the germline and the soma is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes in somatic cells during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells and are then decorated with DNA CpG methylation in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. Currently, most studies have focused on marker genes important for early germ cell development rather than germline genes as a whole given the lack of a curated list of germline-enriched genes. Thus, it is unclear if the mechanism of repression differs for certain classes of germline genes, e.g. meiotic genes versus spermatid differentiation genes. Furthermore, it is currently unknown if ectopic germline gene expression is influenced by chromosomal sex, as previous studies on germline gene repression have been exclusively in males. Sex is particularly pertinent in the case of sexually dimorphic chromatin regulators like KDM5C, which lies on the X chromosome and partially escapes X chromosome inactivation, resulting in 1.2-2-fold higher in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Thus, systematically characterizing KDM5C's role in germline gene repression during early embryogenesis, including its interaction with known silencing mechanisms, will unveil key mechanisms underlying the demarcation between soma and germline identity. These results will provide molecular footholds that can then be exploited to test the contribution of ectopic germline genes on neurodevelopment.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We observed general dysregulation of tissue-enriched genes in both the _Kdm5c_-KO brain and EpiLCs, including testis, liver, muscle, and ovary-enriched genes. The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis genes that are specific to germ cells and not somatic cells of the testis. We found _Kdm5c_-KO EpiLCs aberrantly expressed key drivers of germline identity and meiosis, including _Dazl_ and _Stra8_, while the _Kdm5c_-KO brain primarily expressed germline genes important for late spermatogenesis. Female _Kdm5c_-KO EpiLCs expressed over double the number of germline genes than _Kdm5c_-KO males and germline genes dysregulated in both sexes were expressed more highly in females, demonstrating females have increased sensitivity to germline gene dysregulation. KDM5C was bound to only a subset of germline genes expressed in _Kdm5c_-KO EpiLCs, indicating germline-enriched mRNAs can be aberrantly transcribed through indirect mechanisms. Finally, we found KDM5C promotes the long-term silencing of germline genes in somatic cells by aiding the placement of DNA CpG methylation in EpiLCs. Thus, we propose KDM5C plays a fundamental role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

@@

Previous RNA sequencing (RNA-seq) in the adult hippocampus revealed ectopic expression of some testis genes within the male _Kdm5c_ knockout (-KO) brain[@scandagliaLossKdm5cCauses2017]. Since tissue-enriched genes have not been systematically characterized in the _Kdm5c_-KO brain, if the testis is the only tissue type misexpressed. Therefore, we systematically assessed the expression of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the adult amygdala and hippocampus in mice with constitutive knockout of _Kdm5c_[@vallianatosMutuallySuppressiveRoles2020].
